Investing in infants: the lasting effects of cash transfers to new families

### 2024-07-31 posts/WebPageExample/WebPageExample.qmd
Added WebPageExample.html

---
title: "Investing in infants: the lasting effects of cash transfers to new families"
subtitle: "By Andrew Barr, Jonathan Eggleston, Alexander A. Smith"
author: "Seiro Ito"
format:
  revealjs: 
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    logo: images/quarto.png
    footer: <http://seiroito.github.io/QuartoFiles/BarrEgglestonSmith/BarrEgglestonSmith2022.html>
    theme: sky
include-in-header:
  - text: |
      <style>
      #title-slide
       .title {
        font-size: 2.0em;
      }
      </style>
#width: 1000
bibliography: seiro.bib
resources:
  - BarrEgglestonSmith2022.pdf
# quarto render BarrEgglestonSmith2022.qmd
---
<style>
:root{--r-main-font-size: 36px;}
.reveal h1, .reveal h2, .reveal h3, .reveal h4, .reveal h5 {
              text-transform: none;
      }
.spacer {
    margin-top: 20px;
}
br {
  /* change <br> space length*/
  display: block; /* makes it have a width */
  content: ""; /* clears default height */
  margin-top: 10px; /* change this to whatever height you want it */
}
</style>

@@

::: {.incremental}
* Much affirmative evidence of health and education (in-kind) interventions.  
* Cash intervention impacts are little known.  
   * Food stamps are different from cash [@Hoynes2016; @Baileyetal2023].  
   * Lottery cash distributions are given at older children, results are mixed [@Akeeetal2010; @Cesarinietal2016].  
:::

----

## Research question

::: {.incremental}
* Does providing cash at infancy (< 1 year) have a long run impact on welfare?  
* Use fiscal year cutoff for EITC. Born on:  
   * Dec 31 receives after +1 day  
   * Jan 1 receives after +1 year  
* Examine with tax records + social security numident file (birth day)  
:::

----

@@

::: {.incremental}
* Main tax form used to file a U.S. individual income tax return  
* Every filer  
* Years: 1979, 1984, 1989, 1994-95, 1998-2018  
* Various incomes, dependents (Social Security Numbers)  
:::

----

@@

::: {.incremental}
* Every person born after 1969  
<!-- * Merged with IRS1040 through Social Security Number  -->
* Date of birth, sex, state of birth  
   * Gives: Size of cash transfer in early childhood, family composition  
* $\pm 1$ month of Jan 1, 1982, 1986, 1992  
<!--    * Only these years can be merged with IRS1040 (1979, 1984, 1989)  -->
<!--   * IRS1040 (1979, 1984, 1989) used to predict eligibility for EITC  -->
:::

----

@@

::: {.incremental}
* Students born in 1992-1999 (1993-98 recentred years)  
* Birth dates, academic achievements, behaioral records, free and reduced-price lunch (FRL) eligibility  
   * Income thresholds are similar between EITC and FRL  
      * \$25,600 (FRL) vs. \$27,400 (EITC) in 2000  
      * EITC eligible rate $\simeq$ 75% of all FRL eligible  
* Born within $\pm$ 0928 day windows of Jan 1 (excluding closest 8 days on both sides): 44,992 students  
:::

----

::: {.incremental}
* IRS1040: Parent-dependents linkage  
   * Child's incomes in 23-25, 26-28, 29-31, 32-34 (individual+spouse, 3-year averages)  
* IRS1040+NumidentFile: Date of birth of dependent &rarr; Pre-birth parental incomes &rarr; Pre-birth adjusted gross income (AGI) &rarr; EITC amount  
* Use 1979 incomes (in 1980 tax record) to predict 1980 AGI &rarr; Pre-birth AGI for a Jan 1981 born child  
:::

----

::: {.incremental}
* Predicted AGI using lagged incomes 
   * Attenuation bias  
   * Less concerns of endogeneity (parental ability correlates with AGI, chance of birth date selection)  
* Create an index: Weighted normalised $z$ scores or each outcome $k=\sum\limits_{k}w_{k}\frac{x_{k, treated}-\hat{\mu}_{x_{k, control}}}{\hat{\sigma}_{x_{k, control}}}$ [@Kling2007]

:::

----

@@

----

::: {.incremental}
Strength

@@

1. Some may deliver earlier, confirmed by [@BucklesHungerman2013;  with a c-section, @SchulkindShapiro2014]  
   * No jump in density for firstborns around new year (Fig II)  
   * Balance in covariates (Tab II) 
   * "Donut-hole" RDD: Drop observations within $\pm$ 8 days of Jan 1  
:::

----

2. Discontinuity in other variables  
   * School year begins near Jan 1 in some states, drop these states  
   * Larger tax benefits in later years: Expect larger impacts in later years

----

@@

----

@@

:::: {layout="[40, 40 ]"}

::: {#first-column}
![](FigIII-1.jpg){height=80%}
:::

::: {#second-column}
![](FigIII-2.jpg){height=80%}
:::

::::

@@

----

:::: {layout="[40, 40 ]" layout-valign="bottom"}

::: {#first-column}
![](FigAVa.jpg){height=80%}
:::

::: {#second-column}
![](FigAVb.jpg){height=80%}
:::

::::

@@

----

:::: {.columns}

::: {.column width="70%"}
![](FigV.jpg){height="550"}
:::

::: {.column width="30%"}
* $+319 (23-25), $+456 (26-28), larger at 29+  
* Earnings by 26 is predictive of future earnings [@HaiderSolon2006; @Chettyetal2011]  
:::

::::

----

@@

:::: {.columns}

::: {.column width="65%"}
![](Fig6A.jpg){height="550"}
:::

::: {.column width="35%"}
* Family earnings&uarr; by 4% in 3, 4 years (Fig VI)  
:::

::::

----

:::: {.columns}

::: {.column width="65%"}
![](FigA13.jpg){height="550"}
:::

::: {.column width="35%"}
* Dependents&uarr;  
* Poverty status&darr;  
* Filing tax returns&uarr;  
* Being married  
:::

::::

----

1. Liquidity&uarr;  by 10%-20% during < 12 months old  
   * Hightened expenses and reduced incomes for age < 1 year  
   * Employment continuation through mobility: Car purchases, car parts/repair spending (existing studies)  
   * Intergenerational elasticicty of earning (IGE) $\simeq$ .3. So parental earnings impact ($903.7/58250=1.55$%) $\times$ IGE (1/3)$\simeq 5.16$% or 1/3 of impacts on child's future income is through parental earnings&uarr;

----

2. Stress&darr; (no direct evidence shown in this paper)  
   * Confirmed when additional incomes are provided to single mothers [@SchmidtShoreSheppardWatson2023]  
   * Reduced stress changes the interactions with children [@MilliganStabile2011; @EvansGarthwaite2014]  
   * Marital status stability

----

@@

:::: {.columns}

::: {.column width="55%"}
![](Fig7.jpg){height="550"}
:::

::: {.column width="45%"}
Outcome index&uarr; by .03$\sigma$ per $1000 EITC

@@

See also Tab VIII, Tab A13  
:::

::::
----

:::: {.columns}

::: {.column width="55%"}
![](FigA20.jpg){height="550"}
:::

::: {.column width="45%"}
* Test scores impacts do not fade  
* Suggests human capital channel at play  
* Little gender difference  
   * Due probably to less noise in outcomes than tax filing  
:::

::::
----

::: {.incremental}
* Exam score $\sigma$&uarr; &rArr; $2500&uarr; at age 28 per year [@ChettyFriedmanRockoff2014b]  
* .037$\sigma$ (per \$1000)$\times$ $2500=$92 per year per $1000 spent  
   * 9.2% is pretty high...   
* "(C)ould entirely account for our observed wage effects ... $353 per $1000" (?)  
:::

----

@@

::: {.incremental}
* Long lasting positive impacts of liquidity increase within 12 months after the birth of firstborns  
   * Adult incomes, test scores, behavioral problems, HS graduation  
   * Not evidence for later transfers or non-firstborns  
:::

----

::: {.incremental}
* NPV of tax receipts from added incomes > initial transfer  
* From Sec IV.B: Impact as % earnings per $1000 spent  
   * Food stamps: .3% to 1% 
   * Perry: Under 1%  
   * EITC: 1% to 2%  
* Previous studies on casino wins, lottery, adoption are on wealthier familes, so their impacts are small  
:::

----

>>>>

---
title: "Investing in infants: the lasting effects of cash transfers to new families"
subtitle: "By Andrew Barr, Jonathan Eggleston, Alexander A. Smith"
author: "Seiro Ito"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
    number-depth: 2
    embed-resources: true
    smooth-scroll: true
    anchor-sections: true
bibliography: ../seiro.bib
resources:
  - WebPageExample.pdf
# quarto render WebPageExample.qmd
---
<style>
br {
  /* change <br> space length*/
  display: block; /* makes it have a width */
  content: ""; /* clears default height */
  margin-top: 10px; /* change this to whatever height you want it */
}
</style>

@@

* Much affirmative evidence of health and education (in-kind) interventions.  
* Cash intervention impacts are little known.  
   * Food stamps are different from cash [@Hoynes2016; @Baileyetal2023].  
   * Lottery cash distributions are given at older children, results are mixed [@Akeeetal2010; @Cesarinietal2016].

@@

* Main tax form used to file a U.S. individual income tax return  
* Every filer  
* Years: 1979, 1984, 1989, 1994-95, 1998-2018  
* Various incomes, dependents (Social Security Numbers)

@@

* Every person born after 1969  
<!-- * Merged with IRS1040 through Social Security Number  -->
* Date of birth, sex, state of birth  
   * Gives: Size of cash transfer in early childhood, family composition  
* $\pm 1$ month of Jan 1, 1982, 1986, 1992  
<!--    * Only these years can be merged with IRS1040 (1979, 1984, 1989)  -->
<!--   * IRS1040 (1979, 1984, 1989) used to predict eligibility for EITC  -->

@@

* Students born in 1992-1999 (1993-98 recentred years)  
* Birth dates, academic achievements, behaioral records, free and reduced-price lunch (FRL) eligibility  
   * Income thresholds are similar between EITC and FRL  
      * \$25,600 (FRL) vs. \$27,400 (EITC) in 2000  
      * EITC eligible rate $\simeq$ 75% of all FRL eligible  
* Born within $\pm$ 0928 day windows of Jan 1 (excluding closest 8 days on both sides): 44,992 students

#### How data sets are combined

* IRS1040: Parent-dependents linkage  
   * Child's incomes in 23-25, 26-28, 29-31, 32-34 (individual+spouse, 3-year averages)  
* IRS1040+NumidentFile: Date of birth of dependent &rarr; Pre-birth parental incomes &rarr; Pre-birth adjusted gross income (AGI) &rarr; EITC amount  
* Use 1979 incomes (in 1980 tax record) to predict 1980 AGI &rarr; Pre-birth AGI for a Jan 1981 born child

* Predicted AGI using lagged incomes 
   * Attenuation bias  
   * Less concerns of endogeneity (parental ability correlates with AGI, chance of birth date selection)  
* Create an index: Weighted normalised $z$ scores or each outcome $k=\sum\limits_{k}w_{k}\frac{x_{k, treated}-\hat{\mu}_{x_{k, control}}}{\hat{\sigma}_{x_{k, control}}}$ [@Kling2007]

@@

Strength

@@

1. Some may deliver earlier, confirmed by [@BucklesHungerman2013;  with a c-section, @SchulkindShapiro2014]  
   * No jump in density for firstborns around new year (Fig II)  
   * Balance in covariates (Tab II) 
   * "Donut-hole" RDD: Drop observations within $\pm$ 8 days of Jan 1  
2. Discontinuity in other variables  
   * School year begins near Jan 1 in some states, drop these states  
   * Larger tax benefits in later years: Expect larger impacts in later years

@@

@@

::: {#fig-res1 layout-ncol=2}

![](FigIII-1.jpg){height=80%, #fig3-1}

![](FigIII-2.jpg){height=80%, #fig3-2}

Impacts on child incomes in adulthood
:::

@@

::: {#fig-res2 layout-ncol=2}

![](FigAVa.jpg){width="350"}

![](FigAVb.jpg){width="350"}

Impacts by donut size
:::

@@

::: {.column-margin}
![](FigV.jpg){height="250"}
:::

* $+319 (23-25), $+456 (26-28), larger at 29+  
* Earnings by 26 is predictive of future earnings [@HaiderSolon2006; @Chettyetal2011]

@@

::: {.column-margin}
![](Fig6A.jpg){height="250"}
:::

* Family earnings&uarr; by 4% in 3, 4 years (Fig VI)

::: {.column-margin}
![](FigA13.jpg){height="550"}
:::

* Dependents&uarr;  
* Poverty status&darr;  
* Filing tax returns&uarr;  
* Being married

1. Liquidity&uarr;  by 10%-20% during < 12 months old  
   * Hightened expenses and reduced incomes for age < 1 year  
   * Employment continuation through mobility: Car purchases, car parts/repair spending (existing studies)  
   * Intergenerational elasticicty of earning (IGE) $\simeq$ .3. So parental earnings impact ($903.7/58250=1.55$%) $\times$ IGE (1/3)$\simeq 5.16$% or 1/3 of impacts on child's future income is through parental earnings&uarr;  
2. Stress&darr; (no direct evidence shown in this paper)  
   * Confirmed when additional incomes are provided to single mothers [@SchmidtShoreSheppardWatson2023]  
   * Reduced stress changes the interactions with children [@MilliganStabile2011; @EvansGarthwaite2014]  
   * Marital status stability

@@

::: {.column-margin}
![](Fig7.jpg){height="250"}
:::

Outcome index&uarr; by .03$\sigma$ per $1000 EITC

@@

See also Tab VIII, Tab A13

::: {.column-margin}
![](FigA20.jpg){height="250"}
:::

* Test scores impacts do not fade  
* Suggests human capital channel at play  
* Little gender difference  
   * Due probably to less noise in outcomes than tax filing

* Exam score $\sigma$&uarr; &rArr; $2500&uarr; at age 28 per year [@ChettyFriedmanRockoff2014b]  
* .037$\sigma$ (per \$1000)$\times$ $2500=$92 per year per $1000 spent  
   * 9.2% is pretty high...   
* "(C)ould entirely account for our observed wage effects ... $353 per $1000" (?)

@@

* Long lasting positive impacts of liquidity increase within 12 months after the birth of firstborns  
   * Adult incomes, test scores, behavioral problems, HS graduation  
   * Not evidence for later transfers or non-firstborns  
* NPV of tax receipts from added incomes > initial transfer  
* From Sec IV.B: Impact as % earnings per $1000 spent  
   * Food stamps: .3% to 1% 
   * Perry: Under 1%  
   * EITC: 1% to 2%  
* Previous studies on casino wins, lottery, adoption are on wealthier familes, so their impacts are small
